---
title: "Supplemental Material: Disentangling within- and between-female egg size–number trade-offs in the European earwig"
authors:
  - name: Clint D. Kelly
    affiliations:
      - ref: cdk
    corresponding: true
    email: kelly.clint@uqam.ca
    orcid: 0000-0002-0693-7211
    
affiliations:
  - id: cdk
    name: Département des Sciences biologiques, Université du Québec à Montréal, Montréal, Canada

filters:
  - authors-block
  - iagopinal/figure-table-formatter

crossref:
  tbl-title: Table S
  fig-title: Figure S

format:
  docx:
    toc: false
    reference-doc: word-style-template.docx
bibliography: Bibliography.bib
csl: evolution.csl
---

```{r echo=FALSE}
knitr::read_chunk('scripts/earwig_tradeoff.R')
```

```{r analysis, include=FALSE,echo=FALSE, warning=FALSE, results="hide"}
```

\newpage

## Within–between multivariate model

To partition covariance arising from within-female plasticity from covariance attributable to consistent among-female differences, I applied the within–between decomposition of can de Pol and Wright [@vandepol2009]. Although mean-centering can impose assumptions about underlying biological processes [@westneat2020], these assumptions are appropriate here because the aim was to distinguish plastic allocation within females from persistent differences among females.

Let egg number for female $i$ in brood $j$ be denoted $N_{ij}$. Egg number was decomposed as

$$
N_{ij} = \bar{N}_i + (N_{ij} - \bar{N}_i)
$$ {#eq-decomposition}

The between- and within-female components were defined as

$$
N_i^{B} = \bar{N}_i - \bar{N}
$$ {#eq-between}

$$
N_{ij}^{W} = N_{ij} - \bar{N}_i
$$ {#eq-within}

### Egg number submodel

Egg number was modelled using a negative binomial distribution with a log link:

$$
N_{ij} \sim \mathrm{NegBinomial}(\mu_{ij}, \theta)
$$ {#eq-nb_dist}

$$
\log(\mu_{ij}) =
\beta_{0}^{N}
+ \beta_{1}^{N}\mathrm{Brood}_{ij}
+ u_{0i}^{N}
$$ {#eq-nb_link}

where

$$
u_{0i}^{N} \sim \mathcal{N}(0, \sigma_{N}^{2})
$$ {#eq-nb_random}

### Mean egg size submodel

Mean egg size was modelled as

$$
S_{ij} \sim \mathcal{N}(\mu_{ij}^{S}, \sigma_{S}^{2})
$$ {#eq-gauss_dist}

$$
\mu_{ij}^{S} =
\beta_{0}^{S}
+ \beta_{1}^{S}\mathrm{Brood}_{ij}
+ \beta_{W} N_{ij}^{W}
+ \beta_{B} N_{i}^{B}
+ u_{0i}^{S}
$$ {#eq-gauss_link}

with

$$
u_{0i}^{S} \sim \mathcal{N}(0, \sigma_{S,id}^{2})
$$ {#eq-gauss_random}

### Multivariate structure

The two submodels were fit jointly:

$$
\begin{pmatrix}
u_{0i}^{N} \\
u_{0i}^{S}
\end{pmatrix}
\sim
\mathcal{MVN}
\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\Sigma
\right)
$$ {#eq-mvn}

where

$$
\Sigma =
\begin{pmatrix}
\sigma_{N}^{2} & \sigma_{NS} \\
\sigma_{NS} & \sigma_{S,id}^{2}
\end{pmatrix}
$$ {#eq-sigma_matrix}

The intercept correlation is

$$
\rho_{NS}
=
\frac{\sigma_{NS}}
{\sqrt{\sigma_{N}^{2} \sigma_{S,id}^{2}}}
$$ {#eq-intercept_corr}

## Relationship between model parameters and total phenotypic covariance

Total phenotypic covariance can be partitioned as

$$
\mathrm{Cov}_{\text{total}}
=
\mathrm{Cov}_{\text{within}}
+
\mathrm{Cov}_{\text{among}}
$$ {#eq-cov_partition}

### Within-female covariance

$$
\mathrm{Cov}_{\text{within}}
=
\beta_W \, \mathrm{Var}(N_{ij}^{W})
$$ {#eq-cov_within}

### Among-female covariance

$$
\mathrm{Cov}_{\text{among, slope}}
=
\beta_B \, \mathrm{Var}(N_i^{B})
$$ {#eq-cov_among_slope}

$$
\mathrm{Cov}_{\text{among, intercept}}
=
\sigma_{NS}
$$ {#eq-cov_among_intercept}

Total among-female covariance is therefore

$$
\mathrm{Cov}_{\text{among}}
=
\beta_B \, \mathrm{Var}(N_i^{B})
+
\sigma_{NS}
$$ {#eq-cov_among_total}

### Total phenotypic covariance

Combining components:

$$
\mathrm{Cov}_{\text{total}}
=
\beta_W \, \mathrm{Var}(N_{ij}^{W})
+
\beta_B \, \mathrm{Var}(N_i^{B})
+
\sigma_{NS}
$$ {#eq-cov_total_expanded}

The intercept correlation is

$$
\rho_{NS}
=
\frac{\sigma_{NS}}
{\sqrt{\sigma_N^{2} \sigma_{S,id}^{2}}}
$$ {#eq-rho_definition}

\newpage

::: {.landscape}
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-one
#| tbl-cap: Posterior summaries from the multivariate Bayesian mixed-effects model examining the relationship between egg number and mean egg size. The model jointly estimated a negative binomial model for egg number and a Gaussian model for egg size, including brood effects and partitioning egg number into within- and between-female components. Reported values are posterior medians and 95% credible intervals (CrI). Random effects represent among-female (ID-level) variation, and the correlation term describes the association between female intercepts for egg number and egg size. The dispersion parameter refers to the negative binomial shape parameter.

# -----------------------------
# 1. FIXED EFFECTS (Mean model)
# -----------------------------

fixed_tab <- fit_mv |>
  spread_draws(
    b_meaneggsize_Intercept,
    b_meaneggsize_broodtwo,
    b_meaneggsize_egg_number_within_z,
    b_meaneggsize_egg_number_between_z,
    b_eggnumber_Intercept,
    b_eggnumber_broodtwo
  ) |>
  pivot_longer(cols = -c(.chain, .iteration, .draw),
               names_to = "term",
               values_to = "value") |>
  group_by(term) |>
  median_qi(value, .width = 0.95) |>
  ungroup() |>
  mutate(
    Parameter = recode(term,
      b_meaneggsize_Intercept = "Egg size: Intercept (Brood 1)",
      b_meaneggsize_broodtwo = "Egg size: Brood 2 − Brood 1",
      b_meaneggsize_egg_number_within_z =
        "Egg size: Within-female deviation in egg number (z)",
      b_meaneggsize_egg_number_between_z =
        "Egg size: Between-female deviation in egg number (z)",
      b_eggnumber_Intercept =
        "Egg number: Intercept (Brood 1, log scale)",
      b_eggnumber_broodtwo =
        "Egg number: Brood 2 − Brood 1 (log scale)"
    ),
    Estimate = round(value, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)

# -----------------------------
# 2. RANDOM EFFECTS (Female ID)
# -----------------------------

random_tab <- fit_mv |>
  spread_draws(
    sd_id__meaneggsize_Intercept,
    sd_id__eggnumber_Intercept,
    cor_id__eggnumber_Intercept__meaneggsize_Intercept
  ) |>
  pivot_longer(cols = -c(.chain, .iteration, .draw),
               names_to = "term",
               values_to = "value") |>
  group_by(term) |>
  median_qi(value, .width = 0.95) |>
  ungroup() |>
  mutate(
    Parameter = recode(term,
      sd_id__meaneggsize_Intercept =
        "SD (Egg size | Female ID)",
      sd_id__eggnumber_Intercept =
        "SD (Egg number | Female ID)",
      cor_id__meaneggsize_Intercept__eggnumber_Intercept =
        "Correlation (Egg size, Egg number | Female ID)"
    ),
    Estimate = round(value, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)

# -----------------------------
# 3a. RESIDUAL SD (Egg size)
# -----------------------------

sigma_tab <- fit_mv |>
  spread_draws(sigma_meaneggsize) |>
  median_qi(sigma_meaneggsize, .width = 0.95) |>
  mutate(
    Parameter = "Residual SD (Egg size)",
    Estimate = round(sigma_meaneggsize, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)

# -----------------------------
# 3b. Negative Binomial Shape
# -----------------------------

shape_tab <- fit_mv |>
  spread_draws(shape_eggnumber) |>
  median_qi(shape_eggnumber, .width = 0.95) |>
  mutate(
    Parameter = "Dispersion (Egg number | NegBin shape)",
    Estimate = round(shape_eggnumber, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)

# -----------------------------
# 4. COMBINE ALL TABLES
# -----------------------------

final_tab <- bind_rows(
  fixed_tab,
  random_tab,
  sigma_tab,
  shape_tab
)

flextable(final_tab) |>
  autofit()
```
:::


```{r}
#| echo: false
#| results: asis
#| warning: FALSE
#| label: tbl-two
#| tbl-cap: Posterior medians and 95% credible intervals from a Bayesian mixed model of scaled egg perimeter, including clutch-specific means, clutch-specific female-level random effects, and clutch-specific residual variance.

# -----------------------------
# 1. FIXED EFFECTS (Mean model)
# -----------------------------

fixed_tab <- egg.size.model.alt |>
  spread_draws(
    b_Intercept,
    b_brood_factortwo
  ) |>
  pivot_longer(
    cols = -c(.chain, .iteration, .draw),
    names_to = "term",
    values_to = "value"
  ) |>
  group_by(term) |>
  median_qi(value, .width = 0.95) |>
  ungroup() |>
  mutate(
    Parameter = recode(term,
      b_Intercept = "Mean (Brood 1)",
      b_brood_factortwo = "Brood 2 − Brood 1"
    ),
    Estimate = round(value, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)


# -----------------------------
# 2. RANDOM EFFECTS (ID level)
# -----------------------------

random_tab <- egg.size.model.alt |>
  spread_draws(
    sd_id__brood_factorone,
    sd_id__brood_factortwo,
    cor_id__brood_factorone__brood_factortwo
  ) |>
  pivot_longer(
    cols = -c(.chain, .iteration, .draw),
    names_to = "term",
    values_to = "value"
  ) |>
  group_by(term) |>
  median_qi(value, .width = 0.95) |>
  ungroup() |>
  mutate(
    Parameter = recode(term,
      sd_id__brood_factorone = "SD (Brood 1 | Female ID)",
      sd_id__brood_factortwo = "SD (Brood 2 | Female ID)",
      cor_id__brood_factorone__brood_factortwo =
        "Correlation (Brood 1, Brood 2 | Female ID)"
    ),
    Estimate = round(value, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)


# -----------------------------
# 3. RESIDUAL SD (back-transformed)
# -----------------------------

sigma_tab <- egg.size.model.alt |>
  spread_draws(
    b_sigma_Intercept,
    b_sigma_brood_factortwo
  ) |>
  mutate(
    sigma_brood1 = exp(b_sigma_Intercept),
    sigma_brood2 = exp(b_sigma_Intercept + b_sigma_brood_factortwo)
  ) |>
  select(-.chain, -.iteration, -.draw) |>
  pivot_longer(
    cols = c(sigma_brood1, sigma_brood2),
    names_to = "term",
    values_to = "value"
  ) |>
  group_by(term) |>
  median_qi(value, .width = 0.95) |>
  ungroup() |>
  mutate(
    Parameter = recode(term,
      sigma_brood1 = "Residual SD (Brood 1)",
      sigma_brood2 = "Residual SD (Brood 2)"
    ),
    Estimate = round(value, 3),
    `Lower 95% CrI` = round(.lower, 3),
    `Upper 95% CrI` = round(.upper, 3)
  ) |>
  select(Parameter, Estimate, `Lower 95% CrI`, `Upper 95% CrI`)


# -----------------------------
# 4. COMBINE
# -----------------------------

final_tab <- bind_rows(
  fixed_tab,
  random_tab,
  sigma_tab
)

flextable(final_tab) |>
  autofit()
```


\newpage

# References {.unnumbered}

::: {#refs}
:::
